### 2013-11-12 blad/page2.tex
update

\documentclass[a4paper]{book}

\usepackage[paperwidth=169mm, paperheight=239mm]{geometry}

%\usepackage[latin9]{inputenc}
\usepackage[utf8x]{inputenc}
\usepackage[pdftex]{graphicx}
\usepackage{pdfpages}
\usepackage[british]{babel}
\usepackage{placeins}
\usepackage{dialogue}
\usepackage{fancyvrb}

\usepackage{t1enc}         
\usepackage{times}
\usepackage{url}
\usepackage{enumerate}

%\usepackage{code}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
%\usepackage{lingmacros}
\usepackage{subfig}
\usepackage{fancyhdr}

\pagestyle{plain}
%\pagenumbering{arabic}

%stuff
\usepackage{latexsym}
\usepackage{tikz,pgflibraryshapes}
\usetikzlibrary{shapes,snakes,positioning,matrix}
\usepackage{pgfplots}
\usepackage{booktabs}
\usepackage{enumerate}
%\usepackage{etoolbox}

\usepackage{multibib} 
\newcites{bb}{Bibliography} 
\newcites{ifl}{Bibliography}
\newcites{papp}{Bibliography}
\newcites{exp}{Bibliography}
\newcites{hl}{Bibliography}
\newcites{csort}{Bibliography}
\newcites{emb}{Bibliography}
\newcites{arbb}{Bibliography}
\newcites{t}{Bibliography}
%\newcites[ifl}{Bibliography}

%\patchcmd{\thebibliography}{\section*}{\section}{}{}

\makeatletter
\renewenvironment{thebibliography}[1]
     {\section*{\bibname}%
      %\@mkboth{\MakeUppercase\bibname}{\MakeUppercase\bibname}%
      \list{\@biblabel{\@arabic\c@enumiv}}%
           {\settowidth\labelwidth{\@biblabel{#1}}%
            \leftmargin\labelwidth
            \advance\leftmargin\labelsep
            \@openbib@code
            \usecounter{enumiv}%
            \let\p@enumiv\@empty
            \renewcommand\theenumiv{\@arabic\c@enumiv}}%
      \sloppy
      \clubpenalty4000
      \@clubpenalty \clubpenalty
      \widowpenalty4000%
      \sfcode`\.\@m}
     {\def\@noitemerr
       {\@latex@warning{Empty `thebibliography' environment}}%
      \endlist}
\makeatother

%%

%% \newcommand{\bequ}{\begin{quote}}
%% \newcommand{\enqu}{\end{quote}}
%% \newcommand{\eop}[1]{\mbox{\sl #1}}
%% \newcommand{\sube}[2]{#1_{#2}}
%% \newcommand{\id}[1]{{\it #1}}
%% \newcommand{\kwd}[1]    {\mbox{\textbf{#1}}}
%% \newcommand{\str}[1]{{\id "#1"}}
%% \newcommand{\lemon}{\emph{lemon}}
%% \newcommand{\textt}[1]{\texttt{#1}}
%% \newcommand{\txttt}[1]{\texttt{#1}}

%\renewcommand*{\thesection}{\arabic{section}}

\newcommand{\thesistitle}{Embedded Languages for Data-Parallel Programming}
\newcommand{\dept}{Department of Computer Science and Engineering}
\newcommand{\uni}{Chalmers University of Technology and G\"oteborg University}
\newcommand{\group}{Functional Programming Research Group}

% ---------------------------------------------------------------------------
% Papers 
% ---------------------------------------------------------------------------
\newcommand{\paperA}{Paper A}
\newcommand{\paperATitle}{Simple and Compositional Reification of Monadic Embedded Languages}
\newcommand{\paperB}{Paper B}
\newcommand{\paperBTitle}{Obsidian: A Domain Specific Embedded Language for Parallel Programming of Graphics Processors}
\newcommand{\paperC}{Paper C} 
\newcommand{\paperCTitle}{GPGPU Kernel Implementation and Refinement using Obsidian} 
\newcommand{\paperD}{Paper D}
\newcommand{\paperDTitle}{Expressive Array Constructs in an Embedded GPU Kernel Programming Language}
\newcommand{\paperE}{Paper E}
\newcommand{\paperETitle}{Counting and Occurrence Sort for GPUs using an Embedded Language}
\newcommand{\paperF}{Paper F}
\newcommand{\paperFTitle}{A High-Level Embedded Language for Low-Level GPU Kernel Programming}
\newcommand{\paperG}{Paper G}
\newcommand{\paperGTitle}{Programming Future Parallel Architectures with Haskell and Intel ArBB}
\newcommand{\paperH}{Paper H}
\newcommand{\paperHTitle}{Parallel Programming in Haskell Almost for Free}

% ---------------------------------------------------------------------------
% BEGIN ! 
% ---------------------------------------------------------------------------
\begin{document}

\begin{titlepage}
\begin{centering}
{\sc Thesis for the Degree of Doctor of Philosophy}
\vspace{30ex}

{\LARGE\bf\thesistitle}

\vspace{7ex}

\large Bo Joel Svensson
\vfill
\includegraphics[width=120mm]{./img/ChalmGUtextsvEng}\\[5mm]
\includegraphics[height=4cm]{./img/ChalmGUmarke}\\

\vspace{1cm}
\normalsize
{\sc \dept}\\
{\sc \uni}\\
G\"oteborg, Sweden 2013

\end{centering}
\end{titlepage}

% ---------------------------------------------------------------------------
% Tryckortssida
% ---------------------------------------------------------------------------
%ditt ISBN är:  978-91-7385-939-4
%ditt Ny serie nr:  3620
%ISSN för publ.serien Doktorsavhandlingar vid Chalmers tekniska högskola : 0346-718X

\quad \vfill

{\noindent\large\bf\thesistitle} \\
\noindent Bo Joel Svensson \\
\noindent ISBN 978-91-7385-939-4 \\

\vspace{1cm}

\noindent\copyright {\sc {Bo Joel Svensson}}, 2013 \\ 
\vspace{1cm}

\noindent Doktorsavhandlingar vid Chalmers tekniska högskola \\
\noindent Ny serie nr 3620 \\
\noindent ISSN 0346-718X \\

\vspace{1cm}

\noindent Technical Report 101D\\
\noindent \dept \\
\noindent \group\\

\vspace{1cm}

\noindent \uni \\
\noindent SE--412~96~~G\"oteborg, Sweden\\
\noindent Phone: +46 (0)31--772~1000 \\

\vspace{1cm}

\noindent Printed in Sweden\\
\noindent Chalmers Reproservice\\
\noindent G\"oteborg, Sweden 2013

@@

\clearpage
\pagenumbering{roman}

% ---------------------------------------------------------------------------
% ABSTRACT !!  
% ---------------------------------------------------------------------------
\section*{Abstract}

Computers today are becoming more and more parallel. General purpose processors (CPUs) 
have multiple processing cores and Single Instruction Multiple Data (SIMD) units for 
data-parallelism. Graphics processors (GPUs) bring massive parallelism at the cost of 
being harder to program than CPUs. This thesis applies embedded language methodology 
to data-parallel programming. Two embedded languages are presented, Obsidian for general 
purpose GPU programming and EmbArBB for data-parallel programming across platforms.

@@

\clearpage

% ---------------------------------------------------------------------------
%  ACKS !!! 
% ---------------------------------------------------------------------------
\section*{Acknowledgments}

I would like to thank my supervisor Mary Sheeran. She is a very supportive, motivating and 
understanding supervisor. I also thank my co-supervisor Koen Claessen for all the great advice 
and tips he has provided over the years. I cannot imagine that this would have worked with 
any other supervisors. Thanks to Josef Svenningsson, a late addition to my troop of 
supervisors. Working with Josef has been fun and rewarding.

I have often turned to Emil Axelsson with technical questions, and he has always been
happy to help. For this I am grateful.

I thank Manuel Chakravarty for letting me visit his group at the University of New South Wales 
and Trevor McDonell and Sean Lee for taking care of me during my visit. I look back 
very fondly to my stay in Australia.

For being a great supervisor during my three month internship at Intel, I thank Ryan Newton. 
These where three very intense months away from family but at the same time a very rewarding 
experience.

Thanks to my thesis opponent Stephen Edwards for valuable comments on a draft of this thesis.

I want to thank Erik, Ola, Markus and Viktor for our occasional lunches and GPU discussions. 
I also thank my office mates, Michal and Nicholas, for their company. I thank Eva Axelsson, 
who I have felt most comfortable to turn to for help and information, even when she was not 
the appropriate channel.

Finally I want to thank my family and friends. Thanks to my sisters and parents for their support.
I thank my friends for being understanding during times when I have been very self-absorbed. 
I am very grateful to my wife Andrea for supporting and believing in me and to our daughter Noemi 
for bringing much joy to my breaks from work while writing this thesis; soon she is old enough 
for real LEGO.

\vspace{5mm}
\noindent This research has been funded by the Swedish Foundation for
Strategic Research (which funds the Resource Aware Functional 
Programming (RAW FP) Project) and by the Swedish Research Council.

\clearpage

% ---------------------------------------------------------------------------
%  Guide for Opponent, grading committee
% ---------------------------------------------------------------------------
\section*{Publications}

This thesis includes the following publications:

\begin{enumerate}[A.] 
\item Josef Svenningsson and Bo Joel Svensson. Simple and Compositional Reification of Monadic Embedded Languages, 2013. \emph{18th ACM SIGPLAN International Conference of Functional Programming, ICFP 2013.}
\item Bo Joel Svensson, Mary Sheeran and Koen Claessen. Obsidian: A Domain Specific Embedded Language for General Purpose Parallel Programming of Graphics Processors. In \emph{Proc. of Implementation and Applications of Functional Languages (IFL)}, Lecture Notes in Computer Science, Springer Verlag, March 2009.
\item Bo Joel Svensson, Koen Claessen and Mary Sheeran. GPGPU kernel implementation and refinement using Obsidian. \emph{Practical Aspects of High-level Parallel Programming, PaPP 2010.} Procedia Computer Science.
\item Koen Claessen, Mary Sheeran and Bo Joel Svensson. Expressive Array Constructs in an Embedded GPU Kernel Programming Language. In \emph{Proceedings of the 7th workshop on Declarative aspects and applications of multicore programming}, DAMP '12. ACM. 
\item Josef Svenningsson, Bo Joel Svensson and Mary Sheeran. Efficient Counting Sort Implementations using an Embedded GPU Programming Language. \emph{2nd Workshop on Functional High-Performance Computing, FHPC '13.}
\item Bo Joel Svensson and Mary Sheeran. A High-Level Embedded Language for Low-Level GPU Kernel Programming. {\bf\emph{This work has not yet been published.}} 
\item Bo Joel Svensson and Ryan Newton. Programming Future Parallel Architectures with Haskell and ArBB. \emph{Future Architectural Support for Parallel Programming (FASPP), in conjunction with ISCA '11.}
\item Bo Joel Svensson and Mary Sheeran. Parallel Programming in Haskell Almost for Free: an embedding of Intel's Array Building Blocks. \emph{Proceedings of the 1st ACM SIGPLAN Workshop on Functional High-Performance computing, FHPC '12.}
\end{enumerate}

\clearpage
\subsection*{Personal contributions}

The following describes my personal contributions to the papers included 
in this thesis.

\subsubsection{Paper A}
I intuitively applied the monad reification method as part of the implementation 
of Obsidian. Josef Svenningsson identified its importance and suggested we write 
about it. The paper has two distinct parts. Sections one and two describe the 
method as I used it. And section four makes the compositional aspect of the method 
explicit.

The code and examples in section one and two are my work. The contents of section four was 
developed by Josef Svenningsson and me. For section four we also got very much and valuable help 
from Emil Axelsson.

For the rest of the paper, just substituting my name for Benny and Josef's for Bj\"orn
will give a good indication of our contributions.

\subsubsection{Papers B, C, D and E} 
The Obsidian language has grown out of discussions between Mary Sheeran, Koen Claessen and me. 
When it comes to implementation of the ideas we had in these discussions, I have been responsible. 
Push arrays, first appearing in paper D, were invented by Koen Claessen. However, the 
implementation of them in the setting of Obsidian was done by me. Paper D also has large 
contributions by Mary Sheeran when it comes to examples and implementing them.

Paper E was the first I wrote with Josef Svenningsson, who had been thinking about an interesting
variation of counting sort and wanted to implement it. Together we added atomic operations to 
Obsidian and I modified the code generator to accommodate them. Sections 
1 and 2 in the paper are written by Josef. Sections 3 through 7 are mostly my work, including 
the benchmarking. Section 8 is joint work.
%Me and Josef are equal partners in this paper and got much valuable advice from 
%Mary Sheeran.

\subsubsection{Papers G and H} 
The work behind paper G, work was performed by me 
under supervision of Ryan R. Newton who is also mostly responsible for the writing of the 
paper. Programming and running time experimentation was performed by me, while getting much 
valuable advice and guidance from Ryan.

The work behind paper H was done by me as a hobby project and my aspirations were not 
high. I just wanted to experiment with some embedded language techniques that we had 
been avoiding (had no need for) in Obsidian. Mary discovered what I was doing and applied 
the right amount of pressure, leading to us encompassing much more of ArBB functionality than 
I had in mind. The paper's text is joint work; most examples as well as benchmarking are my work,
but the sparse matrix vector multiplication example is entirely Mary's.

%Intel has unfortunately retired the ArBB project. This leaves EmbArBB without a functioning 
%code generating backend u

% \subsection*{Guide}

\tableofcontents

\cleardoublepage
\clearpage

\pagenumbering{arabic}
\pagestyle{fancy}
\fancyfoot{}
\fancyhead[LO]{}
\fancyhead[RO]{\leftmark}
\renewcommand{\headrulewidth}{0.0pt}
\fancyhead[LE,RO]{\thepage}

% --------------------------------------------------------------------------- %
% --------------------------------------------------------------------------- %
%
%  INTRODUCTION 
%
% --------------------------------------------------------------------------- %
% --------------------------------------------------------------------------- %
\chapter{\thesistitle}

% -------------------------------------------------------------------------
%  INTRO SECTION OF INTRO CHAPTER
% -------------------------------------------------------------------------

\section{Introduction}

This thesis applies embedded language methodology to data-parallel programming of 
massively parallel hardware. An embedded language is a language that is implemented 
as a library for some existing host language. In this case, the functional programming 
language Haskell is used as host language.

Computers today are increasingly parallel. Each new generation of general purpose processors
(CPUs) and of graphics processors (GPUs) brings more parallelism. The question 
of how to program these computers most efficiently, both from a productivity and 
a performance point of view, arises. Explicit parallel programming is hard and automatic 
means of parallelisation, while giving a sequential programming model to the programmer, 
often misses the target when it comes to performance. The book ``Structured Parallel 
Programming'' by Michael McCool et al argues the need for explicit parallel programming 
models~\citet{STRUCTURED}.

The embedded language approach to programming language implementation is beneficial, 
since we are not entirely sure of how to program current and future highly parallel 
computers. Implementing an embedded language allows us more rapid prototyping of ideas. 
We can experiment with what abstractions to expose to the programmer, without much 
effort. There are many other benefits of embedded languages: making use 
of the host language's library functions (if they are general enough), the host language 
can be used as a powerful macro language for the embedded language, providing a more 
leveled learning curve to programmers already familiar with the host language. 
For programmers who are already proficient in the host language, learning an embedded 
language is more like picking up a new library.

This thesis focuses on data-parallelism, as opposed to task or control parallelism.
Data-parallelism is applicable where the input data can be divided into chunks
and operated upon independently. Data-parallelism is scalable; the potential 
for parallelism increases with the problem size. Two different embedded languages 
for data-parallel programming are presented in this thesis: Obsidian, for 
general purpose computations on GPUs and EmbArBB for parallel programming across 
platforms, such as CPUs, GPUs and other accelerators.

Graphics Processors trade programmer convenience for increased parallelism. 
For example, where CPUs dedicate large portions of chip area to caches, branch 
prediction and instruction level parallelism (ILP), GPUs put additional cores. 
This comes at a price of increased programmer effort. GPUs have programmer managed 
shared memories, conditionals are troublesome from a performance 
perspective and memory access patterns can make or break the performance of an 
application. With Obsidian, we try to find the right level of abstraction for 
programming GPUs. We try to strike a balance between low-level control of details 
that influence performance and high-level abstractions to increase productivity and 
make programs more terse. The end goal with Obsidian is a language for writing 
high performance GPU programs that compare well to code written by CUDA~\citet{CUDA}  
(a C-dialect for GPU programming) experts, provided that both programmers have 
similar GPU expertise. The Obsidian programmer would however write programs in a more 
composable manner. Obsidian programs are compiled into CUDA.

Obsidian is geared towards the programmer who is interested in GPU hardware. Making 
good use of Obsidian requires that the programmer is willing to make an effort 
to learn about GPUs and how to program them efficiently. However, another case 
is a programmer who does not want to invest that effort but wants to write a high-level 
parallel program and have it run reasonably well on different parallel architectures. 
In this case, a language with even higher level of abstraction is preferable. It then 
becomes the job of a more advanced compiler to turn that into code for the targeted 
platforms. The EmbArBB language targets this programmer. EmbArBB exposes a set of high-level 
operations on arrays, such as {\tt map}, {\tt reduce} and {\tt scan}. The programmer 
has no power over the parallel strategy that should be used to implement these 
operations but instead relies on a JIT-compiler (Just In Time-compiler). EmbArBB is a 
Haskell embedding of Intel ArBB~\citet{ARBB2011} that was released as an embedded language 
in C++. EmbArBB relies entirely on the ArBB system for code generation and optimisation.

% -------------------------------------------------------------------------
% Obsidian !
% -------------------------------------------------------------------------

\section {Obsidian: An embedded language for GPU kernel implementation}

In 2006, NVIDIA released their first Compute Unified Device Architecture (CUDA) 
GPU. With CUDA, NVIDIA also made it a lot easier for programmers to use GPUs
for general purpose programming~\citet{wwwcuda}. A dialect of C, called CUDA C, was released 
specifically to simplify the implementation of general purpose computations on GPUs.

Before CUDA, programmers still tried (very ingeniously) to make use of GPUs for 
non-graphics applications via the graphics API. The desired computation was expressed 
in graphics terms and the result computed through an act of rendering. Of course, since the 
available abstractions targeted the graphics domain, this added extra
burden on the programmer. CUDA was NVIDIA's first response to this problem. CUDA 
is a dialect of C, extended with a parallel programming model suitable for GPUs.

Having a language specifically made for general purpose programming on the GPU, such as 
CUDA, is a big step up from earlier. But CUDA still offers a very low-level interface to 
programming the GPU, which is preferred in a situation where the programmer needs to 
think about hardware specific details to get the best performance. CUDA provides this, 
but at the price that it is hard to build upon, abstract and reuse prior programming 
effort. This is where we are trying to go one step further with Obsidian.

% -------------------------------------------------------------------------
% GPU Hardware 
% -------------------------------------------------------------------------

\subsection{GPU hardware}

Before going into programming of GPUs, it is important to have some background. The 
GPU programming model exposed by CUDA very much mirrors the underlying hardware. Some 
of the details that make GPU programming hard are more apparent when looking at the 
underlying hardware.

A CUDA GPU is built around a single kind of processor (as opposed to the different 
kinds of processors found in earlier GPUs). The processors in the GPU (called MPs, 
MultiProcessors) all contain a number of cores called SPs (Streaming Processors). 
Each MP also contains local memory, called shared memory since it can be accessed 
by all of the SPs in that MP. The number of MPs varies over the available GPUs; cheaper 
GPUs have as few as one MP, and as you go up in price the number of MPs increases.

Each MP of the GPU can manage a large number of threads; on today's GPUs up to 
2048 threads can run on a single MP. The GPU schedules threads in groups of 32, called 
{\em Warps}, that are executed in lock-step (SIMD style execution). Threads are 
also divided into {\em Blocks}; the threads within a block can communicate using the 
shared memory. The maximum block size is currently 1024 threads. Threads within 
a warp can communicate via the shared memory without using any synchronisation 
primitive. However, if communication takes place across warps synchronisation is 
necessary. A barrier synchronisation mechanism exists to ensure that all threads within a block 
have reached the same position in the code. Blocks are also grouped into a grid that 
is the collection of blocks executing the same program.

% -------------------------------------------------------------------------
%  GPU Programming
% -------------------------------------------------------------------------

\subsection{GPU programming} 
\label{sec:GPUPROGRAMMING}

The CUDA programming model reflects the hierarchical architecture of the GPU. The 
programmer writes what are called \emph{kernels}; in CUDA jargon, a kernel refers to a sequential 
C program that is parameterised on a thread's identity. The kernel program is executed 
in parallel across all the threads of a block and many such blocks of threads can execute 
in parallel on the available MPs. Kernels may use shared memory that is local to a block; 
the programmer needs to decompose the algorithm and data in such a way that communication 
will be local to the blocks.

The program below is a very simple example of a CUDA kernel that computes elementwise sums.

\begin{small}
\begin{Verbatim}[samepage=true]
__global__ void vecAdd(int32_t *i1, int32_t *i2, int32_t *r) {

  unsigned int gid = blockIdx.x * 
                     blockDim.x + 
                     threadIdx.x;

  r[gid] = i1[gid] + i2[gid];

}

\end{Verbatim}
\end{small}

\noindent The kernel takes two input arrays, {\tt i1} and {\tt i2}; the result is stored 
in {\tt r}. Note how the kernel makes use of three variables, {\tt threadIdx.x}, 
{\tt blockIdx.x} and {\tt gridDim.x}. These variables identify a thread uniquely within 
a grid. Here they are used to obtain an index into the provided arrays. The dimensions of 
the blocks and of the grid are set by a controlling program upon launch of the kernel. This 
controlling program is run by the CPU of the computer housing the GPU. The kernel is launched 
using the following syntax:

\begin{small} 
\begin{Verbatim}[samepage=true]
vecAdd<<<1,32,0>>>(d_i1,d_i2,d_r);
\end{Verbatim}
\end{small}

\noindent This line of code launches the kernel on one block of 32 threads. The zero 
indicates that the kernel uses no shared memory. If the kernel did use shared memory, the 
amount needed should have been specified there. Figure~\ref{fig:CUDACODE1} shows the complete 
sequence of commands for a minimal example using the kernel specified above.

% -------------------------------------------------------------------------
% CUDACODE1 FIGURE
% -------------------------------------------------------------------------

\begin{figure} 
\begin{small}
\begin{Verbatim}[samepage=true]
int main(void) {

  int32_t h_i1[32], h_i2[32], h_r[32];

  int32_t *d_i1, *d_i2, *d_r;

  for (int i = 0; i < 32; ++i) { 
    h_i1[i] = i;
    h_i2[i] = 32 - i;
  }

  // Allocate device memory
  cudaMalloc((void**)&d_i1,32*sizeof(int32_t));
  cudaMalloc((void**)&d_i2,32*sizeof(int32_t));
  cudaMalloc((void**)&d_r,32*sizeof(int32_t));

  // Copy inputs to device
  cudaMemcpy(d_i1,
             h_i1,
             32*sizeof(int32_t),
             cudaMemcpyHostToDevice);
  cudaMemcpy(d_i2,
             h_i2,
             32*sizeof(int32_t),
             cudaMemcpyHostToDevice);

  // Launch the kernel
  vecAdd<<<1,32,0>>>(d_i1,d_i2,d_r);

  // Copy result to host
  cudaMemcpy(h_r,
             d_r,
             32*sizeof(int32_t),
             cudaMemcpyDeviceToHost);

  for (int i = 0; i < 32; ++i) { 
    printf("%d ",h_r[i]);
  }

  return 0;
}
\end{Verbatim} 
\end{small}

\caption{Host code for launching a simple CUDA kernel on the GPU.}
\label{fig:CUDACODE1}

\end{figure}

% -------------------------------------------------------------------------
% 
% -------------------------------------------------------------------------

\noindent In Obsidian, the same kernel (the local computation) is implemented as follows:

\begin{small}
\begin{Verbatim}[samepage=true]
vecAdd :: Num a => SPull a -> SPull a -> SPull a
vecAdd = zipWith (+)
\end{Verbatim}
\end{small}

\noindent The {\tt vecAdd} program takes two {\tt SPull} input arrays and gives one 
as result. The Obsidian {\tt vecAdd} is more general than the CUDA version. From it 
we can generate CUDA kernels for any element type that supports the {\tt (+)} operation. 
In Haskell we know that types that belong to the {\tt Num} typeclass has a {\tt (+)} operation.

There are some important differences between CUDA and Obsidian programs. One is that with Obsidian 
we generate kernels for fixed size local computations. This is to avoid doing out-of-bounds checks 
which would mean conditionals in the code. Conditionals are problematic from a performance 
point of view. Another difference is that in Obsidian the programmer specifies the entire 
computation, including how it is mapped over the blocks of the GPU. This is done implicitly 
in CUDA, based on the indexing arithmetic used in the kernel and the number of blocks and 
threads provided at kernel launch. This means that the Obsidian programmer needs to 
write a program that performs this mapping:

\begin{small} 
\begin{Verbatim}[samepage=true]
vecAddG :: Num a => DPull a -> DPull a -> DPush Grid a
vecAddG i1 i2  =
  pConcat $
  fmap push $ 
  zipWith vecAdd
          (splitUp 32 i1)
          (splitUp 32 i2)  
\end{Verbatim}
\end{small}

The {\tt vecAddG} program specifies a computation on large arrays; the large input arrays 
are split up into pieces that are operated upon by the local computation, {\tt vecAdd}, 
specified above. Each input array passed to this function must have length a multiple of 32.
Since the {\tt splitUp} operation splits the arrays into chunks of 32 elements, the body 
of the generated kernel will be constructed specifically for this size. {\tt zipWith} is 
used to map the local computation onto each chunk of the input arrays. The result is an 
array of arrays that is flattened with the {\tt pConcat} function. The CUDA code below 
is generated from the {\tt vecAddG} program:

\begin{small}
\begin{Verbatim}[samepage=true] 
extern "C" __global__ void vecAdd(int32_t* input0, uint32_t n0,
                                  int32_t* input1, uint32_t n1,
                                  int32_t* output2)
{
    uint32_t tid = threadIdx.x;

    if (blockIdx.x < min(n0 / 32U, n1 / 32U)) {
        output2[blockIdx.x * 32U + tid] = 
            input0[blockIdx.x * 32U + tid] +
            input1[blockIdx.x * 32U + tid];
    }
}
\end{Verbatim}
\end{small}

There are some superficial differences between the code generated using Obsidian and the 
handwritten example above. One such difference is that the index computation is not shared. 
Obsidian relies on the CUDA compiler to discover such sharing via a Common Subexpression 
Elimination (CSE) optimisation pass. Another difference is that Obsidian has inserted a 
conditional that only executes the body for certain blocks. Discussion of the causes for 
that conditional is postponed until section~\ref{sec:OBSARRAYS}. A real 
difference is that the kernel is generated for a fixed size; here there is a constant (32) 
where in the handwritten code we used {\tt blockDim.x}. If those differences are set 
aside, the two kernels are identical. Figure~\ref{fig:OBSIDIANHOST1} shows how the 
{\tt vecAddG} kernel can be launched from within Haskell, using Obsidian's rudimentary 
support for launching GPU computations.

An alternative way to implement {\tt vecAddG} is to inline the local computation 
directly. This makes the entire program look like this.

\begin{small}
\begin{Verbatim}[samepage=true]
vecAddG :: Num a => DPull a -> DPull a -> DPush Grid a
vecAddG i1 i2  =
  pConcat $
  fmap push $
  zipWith (zipWith (+)) 
          (splitUp 32 i1)
          (splitUp 32 i2)
\end{Verbatim} 
\end{small}

\noindent The generated code looks exactly as before.

% -------------------------------------------------------------------------
% OBSIDIANHOST1 FIGURE
% -------------------------------------------------------------------------

\begin{figure} 
\begin{small}
\begin{Verbatim}[samepage=true]
performVecAdd =
  withCUDA $
  do
    kern <- capture 32 (vecAddG :: DPull EInt32
                                 -> DPull EInt32
                                 -> DPush Grid EInt32)

    useVector (V.fromList [0..32::Int32]) $ \i1 ->
      useVector (V.fromList [0..32::Int32]) $ \i2 -> 
        allocaVector 32 $ \(o :: CUDAVector Int32) ->
        do
          fill o 0
          o <== (1,kern) <> i1 <> i2

          r <- peekCUDAVector o
          lift $ putStrLn $ show r
\end{Verbatim} 
\end{small}

\caption{Host code for launching a simple Obsidian kernel on the GPU. \\ 
 {\tt capture} compiles an Obsidian program into CUDA. At this point the 
  types of the Obsidian program need to be made concrete. The number, 32, used as 
 an argument to {\tt capture} specifies that generated code should be specialised 
 for 32 threads per block. }
\label{fig:OBSIDIANHOST1}
\end{figure}

\pagebreak

\noindent A CUDA kernel that uses shared memory takes the form:

\begin{small}
\begin{Verbatim}[samepage=true]
__global__ void kernel(int32_t *i, int32_t *r) {

  extern __shared__ int32_t sm[];

  unsigned int tid = threadIdx.x; 
  unsigned int gid = blockIdx.x * 
                     blockDim.x + 
                     threadIdx.x;

  sm[tid] = i[gid]; 
  __syncthreads();

  /* Compute on sm */

  __syncthreads();
  r[gid] = sm[tid];

}
\end{Verbatim} 
\end{small}

\noindent A portion of the input array is stored into shared memory. A 
barrier synchronisation is used to ensure that each thread within the block 
has stored its value into shared memory (only necessary if threads will 
communicate across warp boundaries). After computing on the data locally, 
it is written back into global memory. This kernel sets up two aliases for 
thread identities. A thread's global identity is called {\tt gid} and its local 
identity is called {\tt tid}. The global ID is used to index into the large global 
array; indexing into local shared memory is done using the local ID.

In Obsidian, local computations can also use shared memory. The Obsidian arrays 
we have seen in the examples, {\tt Pull} and {\tt Push} arrays do not directly 
correspond to actual data in memory. Pull and push arrays rather represent ways 
to compute arrays. More information about these array representations is available 
in section~\ref{sec:OBSARRAYS}. The Obsidian programmer can store intermediate 
arrays in shared memory by using a {\tt force} operation. 
%The 
%{\tt force} operation computes an array and stores it into shared memory.

The following kernel multiplies each element in an array by two and then adds one:

\begin{small}
\begin{Verbatim}[samepage=true] 
kernel :: Num a => SPull a -> SPull a
kernel = fmap (+1) . fmap (*2)
\end{Verbatim}
\end{small}

\noindent This kernel computes its result without any storage of intermediate results;
the code is equivalent to \verb!fmap ((+1).(*2))!. 
In the generated CUDA code below, it can be seen that the (*2) and (+1) operations 
are applied without any intermediate storage.

\begin{small}
\begin{Verbatim}[samepage=true]
extern "C" __global__ void kernel(int32_t* input0, uint32_t n0,
                                  int32_t* output1)
{
    uint32_t tid = threadIdx.x;

    if (blockIdx.x < n0 / 32U) {
        output1[blockIdx.x * 32U + tid] = 
          input0[blockIdx.x * 32U + tid] * 2 + 1;
    }
}
\end{Verbatim}
\end{small}

Storing an intermediate result between the two operations can be accomplished by 
changing the kernel as follows.

\begin{small}
\begin{Verbatim}[samepage=true]
kernelF :: (MemoryOps a, Num a)
         => SPull a -> BProgram (SPull a)
kernelF = liftM (fmap (+1)) . force . fmap (*2)
\end{Verbatim} 
\end{small}

This change introduced some noise. The reason for this is that Obsidian programs that 
use memory are monadic. The {\tt BProgram} in the result type is a monad. Another difference 
is that we need to be able to write the array elements to memory; in Obsidian this means 
that the element type must be in the {\tt MemoryOps} class.

The code below is generated using the {\tt kernelF} program:

\begin{small} 
\begin{Verbatim}[samepage=true]
extern "C" __global__ void kernel(int32_t* input0, uint32_t n0,
                                  int32_t* output1)
{
    extern __shared__ uint8_t sbase[];
    uint32_t tid = threadIdx.x;

    if (blockIdx.x < n0 / 32U) {
        ((int32_t*) sbase)[tid] = 
           input0[blockIdx.x * 32U + tid] * 2;
        __syncthreads();
        output1[blockIdx.x * 32U + tid] = 
           ((int32_t*) sbase)[tid] + 1;
    }
}
\end{Verbatim}
\end{small}

\noindent This code uses a shared memory array called {\tt sbase}. There 
is also a call to \\ {\tt \_\_syncthreads} to ensure that all the writes have 
completed before proceeding. In this case the synchronisation is not really
necessary since all threads belong to the same warp. There is an {\tt unsafeforce} 
operation that naively removes synchronisations if the array being forced is short 
enough. More advanced analysis could also be performed to remove synchronisations 
whenever communication is entirely warp local. We do not apply that optimisation, 
but work on warp level computations is in progress and would put this ability 
in the hands of the programmer.

\FloatBarrier 
\subsection{Writing efficient GPU Kernels} 
\label{sec:efficient}

Writing efficient GPU kernels is hard. There are many details that the CUDA 
programmer must be aware of in order to make optimal use of the GPU. 
The NVIDIA ``CUDA C Best Practices Guide''~\citet{BestPrac}, is a good source 
to learn more about these details. Here is a list, with explanations of some 
of the practices that the guide deems most important. The list also contains 
information about how these best practices influence our work on Obsidian.

\begin{itemize}

\item {\bf Minimise data transfers between host and device:} The bandwidth between device memory 
and GPU is much higher than the bandwidth between host memory and the GPU device via the PCIe bus.
This means that, first of all, one must evaluate if the computation to be performed is significant 
enough to warrant a transfer to the GPU. If the computation is not arithmetic intensive, it may be 
better to do the work on the CPU. However, if the data is already in device memory and 
the operation that is about to be performed on it is more efficiently handled by the CPU, it 
may still be better to let the GPU do the work. In other words, compute near the data unless 
transferring it pays off.

Data transfers between the host memory and device (GPU) memory is out of Obsidian's scope. 
If Obsidian is used to generate kernels for a computation, the programmer has already decided that 
a GPU should be used. However, even after making the decision to use the GPU the programmer 
may need to consider how to transfer the data. In some cases it is possible to overlap 
computation on the GPU and data transfer. These kinds of optimisations are considered 
when writing the C program that launches the CUDA kernels. Obsidian does have rudimentary 
support for launching kernels from within Haskell but currently we consider this mostly a
tool for quickly being able to run a kernel and see what it does, not as a means to implement 
applications. This part of Obsidian is future work.

\item {\bf Ensure that global memory accesses are coalesced:} Certain access patterns into 
GPU device memory can be {\em coalesced} (combined into few memory transactions). Unfortunately 
what these access patterns are varies somewhat depending on the GPU. For a typical model of 
GPU, the rule is that concurrent memory accesses (by threads within a warp) will be coalesced 
to as many transactions as cache lines touched (128 Byte l1 cache lines). This means that strided 
accesses from within a warp are not favoured.

Inputs to an Obsidian program are pull arrays. Pull arrays can be arbitrarily permuted; 
the permutation used will decide the access pattern. In section~\ref{sec:OBSARRAYS}, there 
is more information about pull arrays. If global memory accesses can 
be coalesced or not partly also depends on the algorithm being implemented. The next paragraph 
suggests a trick for coalescing memory accesses, in some cases, even if the algorithm is 
unsuitable. Then it becomes a trade between bad memory access pattern and using more 
shared memory.

>>>>

@@

\section*{Abstract}

\begin{small}
Computers today are becoming more and more parallel. General purpose processors (CPUs) 
have multiple processing cores and Single Instruction Multiple Data (SIMD) units for 
data-parallelism. Graphics processors (GPUs) bring massive parallelism at the cost of 
being harder to program than CPUs. This thesis applies embedded language methodology 
to data-parallel programming. Two embedded languages are presented, Obsidian for general 
purpose GPU programming and EmbArBB for data-parallel programming across platforms.

@@

\end{small}
\newpage
